---
title: 'Using funneljoin to analyze #rstats twitter'
author: Emily Robinson
date: '2019-08-05'
slug: using-funneljoin-to-analyze-rstats-twitter
categories: []
tags: []
keywords:
  - tech
---

In my [previous post](https://hookedondata.org/introducing-the-funneljoin-package/), I used the `after_join()` function from my package [funneljoin](https://github.com/robinsones/funneljoin) to analyze Stack Overflow questions and answers that used the tag R. In this post, I'll introduce funneljoin's `funnel_start()` and `funnel_step()` functions to look at tweets tagged with `#rstats`. This data comes from the [R for Data Science](https://www.rfordatasci.com/) [tidytuesday](https://github.com/rfordatascience/tidytuesday) initiative; you can download it [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-01-01/rstats_tweets.rds). 

```{r message = FALSE}
library(tidyverse)
# devtools::install_github("robinsones/funneljoin")
library(funneljoin)
tweets <- readRDS("rstats_tweets.rds")
```

Let's take a look at the dataset: 

```{r}
tweets
```

The first thing we notice is the large number of columns (88!) - we're certainly not going to need them all. I'll start by getting just a subset. 

```{r}
tweets %>%
  select(created_at, screen_name, text, is_quote, favorite_count, retweet_count, hashtags, urls_expanded_url, mentions_screen_name, followers_count)
```

Who tweets with the r hashtag and then with another?

```{r}
tweets %>%
  mutate(nb_hashtags = map_dbl(hashtags, ~ length(.)))
```

